Gene expression responses to diet quality and viral infection in Apis mellifera

### 2018-09-16 Manuscript/Text/manuscript.tex
LME and pairwise BH

As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.30e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 1.00e-05) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Tukey HSD, p $<$ 1.31e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Tukey HSD, p $<$ 4.78e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 1.96e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p $<$ 5.65e-03). However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Tukey HSD, p = 1.11e-01). Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not do so (Figure 2).

@@

While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies is even greater than this as we used naturally mated honey bees from 15 different colonies. We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. It is our hope that this aspect of our study may shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.

>>>>

As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Namely, bees fed Rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed Chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02). Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not do so (Figure 2).

@@

While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies. We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. It is our hope that this aspect of our study may shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees.
